A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-14 sections/Overview.tex
revised 2.3

%-------------------------------------------------------------------------------
\subsection{Small-Step Unification}
% Mention some algorithms for unification for 
% dependent types that use a small-step approach. Credit them later for some ideas that 
% we also employ.
There are literals that make use of small-step unification for dependently typed
inference and reconstruction algorithms~\cite{Reed2009,Abel2011higher}.
These approaches collect a list of constraints and process one at a time.
Similar to DK's algorithm, unification variables are used to represent the types to guess.
In order to solve unification variables to correctly scoped types
(typically constrained by the context when the variable is introduced),
the context information should be kept with the variables.
Judgments that represent unifications of terms might also require such a context
to keep the constraint collection well-formed.

As an example, Abel et al. defines their constraint $K$ by
\begin{gather*}
\begin{aligned}
K &::= {\top} \mid {\bot} &&\text{Trivial constraint and inconsistency.}\\
    & \ \mid \  \Psi\vdash M = N : C &&\text{Unify term $M$ with $N$.}\\
    & \ \mid \  \Psi\mid R:A \vdash E = E' &&\text{Unify evaluation context $E$ with $E'$.}\\
    & \ \mid \  \Psi\vdash u\leftarrow M: C &&\text{Solution for $u$ (metavariable) found.}
\end{aligned}
\end{gather*}
where the unification constraints and metavariable solutions are bound by a proper context.
A unification problem in their system is described by $\Delta \Vdash \overline{K}$,
where $\Delta$ is a collection of metavariables with their defined scopes.
This approach clearly states the scroping of each unification problem,
therefore rules out invalid instantiations to metavariables.

However, such ``duplicated contexts'' are not ideal for our formalization.
Following DK's algorithm, an existential variable $\al$ defined in the context
could be decomposed into a function type $\al[1] \to \al[2]$,
so the declaration of $\al$ should be replaced by two declarations $\al[1], \al[2]$.
Such operation causes all the context that contains $\al$ to change,
which brings difficulty on the synchronization of multiple contexts,
as variable declarations are not centralized.

We find a way of encoding multiple contexts by ``compressing'' them into a single worklist.
As our type inference rules applied to a judgment,
we typically keep the context or add variable declarations to the old context
before analysing its sub-judgment(s).
This enables us to write the judgment worklist in such a form:
$$\Gm, \{\text{variable declarations}\}, \{\text{judgment}\},
\{\text{variable declarations}\}, \{\text{judgment}\}, \ldots .$$
Typically, we simplify the right-most judgment and push back smaller tasks,
where the old ``context'' automatically gets inherited.
When some rules solve or partially solve an existential variable,
we could easily propagate the solution to all the judgments,
and safely modify its declaration, such as remove from the worklist.
New judgment(s) may also introduce new \emph{local} variables by declaring
that variable right before the new judgment(s).
After the new judgments are satisfied, these local variables are properly recycled.

% \jimmy{TODO explain the duplicated context with some examples (their judgment form)}
% Some issues to point out: Duplicated contexts (rather than shared contexts), which make 
% the formalization harder since requires ``synchronizing'' things ...

%-------------------------------------------------------------------------------
\subsection{Worklist Algorithm for Subtyping}

>>>>

%-------------------------------------------------------------------------------
\subsection{Judgment Lists}

To avoid the problem of incrementally applying a substitution to remaining
tasks, we can find inspiration in the formulation of constraint solving
algorithms. For instance, the well-known unification
algorithm by \citet{unification} decomposes the problem of unifying two terms $s \stackrel{.}{=} t$ into a set
of related unification problems between pairs of terms $s_i \stackrel{.}{=} t_i)$. These smaller
problems are not tackled independently, but kept together in a set $S$. 
The algorithm itself is typically formulated as a small-step-style state
transition system $S \rightarrowtail S'$ that rewrites the set of unification
problems until it is in solved form or until a contradiction has been found.
For instance, the variable elimination rule is written as:
\[
   x \stackrel{.}{=} t, S  ~~\rightarrowtail~~  x \stackrel{.}{=} t, [t/x]S   \qquad\qquad{if}~x \not\in t~\text{and}~{x \in S}
\]
Because the whole set is explicitly available, the variable $x$ can be
simultaneously substituted everywhere.

In the above standard unification problem, all variables are implicitly in the same
global scope. In contrast, recent unification algorithms for dependently-typed languages
are more explicit about scopes. For instance, \citet{Reed2009} represents a unification
problem as $\Delta \vdash P$ where $P$ is a set of equations to be solved and $\Delta$ is
a (modal) context. \citet{Abel2011higher} even use multiple contexts within a unification problem.
Such a problem is denoted $\Delta \Vdash \mathcal{K}$ where the meta-context
$\Delta$ contains all the typings of meta-variables in the constraint set
$\mathcal{K}$. The latter consists of constraints like $\Psi \vdash M = N : C$
that are equipped with their individual context $\Psi$. While acurrately tracking
the scoping of regular and meta-variables, this approach is not ideal because it
repeatedly copies contexts when decomposing a unification problem, and later has to 
substitute solutions into these copies.

% %-------------------------------------------------------------------------------
% \subsection{Small-Step Unification}
% % Mention some algorithms for unification for 
% % dependent types that use a small-step approach. Credit them later for some ideas that 
% % we also employ.
% There are literals that make use of small-step unification for dependently typed
% inference and reconstruction algorithms~\cite{Reed2009,Abel2011higher}.
% These approaches collect a list of constraints and process one at a time.
% Similar to DK's algorithm, unification variables are used to represent the types to guess.
% In order to solve unification variables to correctly scoped types
% (typically constrained by the context when the variable is introduced),
% the context information should be kept with the variables.
% Judgments that represent unifications of terms might also require such a context
% to keep the constraint collection well-formed.
% 
% As an example, Abel et al. defines their constraint $K$ by
% \begin{gather*}
% \begin{aligned}
% K &::= {\top} \mid {\bot} &&\text{Trivial constraint and inconsistency.}\\
%     & \ \mid \  \Psi\vdash M = N : C &&\text{Unify term $M$ with $N$.}\\
%     & \ \mid \  \Psi\mid R:A \vdash E = E' &&\text{Unify evaluation context $E$ with $E'$.}\\
%     & \ \mid \  \Psi\vdash u\leftarrow M: C &&\text{Solution for $u$ (metavariable) found.}
% \end{aligned}
% \end{gather*}
% where the unification constraints and metavariable solutions are bound by a proper context.
% A unification problem in their system is described by $\Delta \Vdash \overline{K}$,
% where $\Delta$ is a collection of metavariables with their defined scopes.
% This approach clearly states the scroping of each unification problem,
% therefore rules out invalid instantiations to metavariables.
% 
% However, such ``duplicated contexts'' are not ideal for our formalization.
% Following DK's algorithm, an existential variable $\al$ defined in the context
% could be decomposed into a function type $\al[1] \to \al[2]$,
% so the declaration of $\al$ should be replaced by two declarations $\al[1], \al[2]$.
% Such operation causes all the context that contains $\al$ to change,
% which brings difficulty on the synchronization of multiple contexts,
% as variable declarations are not centralized.
% 
% We find a way of encoding multiple contexts by ``compressing'' them into a single worklist.
% As our type inference rules applied to a judgment,
% we typically keep the context or add variable declarations to the old context
% before analysing its sub-judgment(s).
% This enables us to write the judgment worklist in such a form:
% $$\Gm, \{\text{variable declarations}\}, \{\text{judgment}\},
% \{\text{variable declarations}\}, \{\text{judgment}\}, \ldots .$$
% Typically, we simplify the right-most judgment and push back smaller tasks,
% where the old ``context'' automatically gets inherited.
% When some rules solve or partially solve an existential variable,
% we could easily propagate the solution to all the judgments,
% and safely modify its declaration, such as remove from the worklist.
% New judgment(s) may also introduce new \emph{local} variables by declaring
% that variable right before the new judgment(s).
% After the new judgments are satisfied, these local variables are properly recycled.
% 
% % \jimmy{TODO explain the duplicated context with some examples (their judgment form)}
% % Some issues to point out: Duplicated contexts (rather than shared contexts), which make 
% % the formalization harder since requires ``synchronizing'' things ...
% 
%-------------------------------------------------------------------------------
\subsection{Single-Context Worklist Algorithm for Subtyping}
